On Devon Allen's Disqualification at the 2022 World Track and Field Championships

### 2023-07-20 Manuscript/manuscript.tex
Started to clean up the discussion section

The dataset of the four United States athletes 
is too small to perform any meaningful statistical analysis, but
the data was expanded to include reaction times for athletes who competed at 
national competitions (many of which were held from May-July of 2022). Once the 
data was collected, each athlete was designated as a cluster and the stage of
competition (national or international) was specified for each competition 
within the cluster (athlete).  We chose to include preliminary rounds in this
section as well as in \ref{sec:data2019} in order to increase the size of our
data.  Without data from the preliminary rounds, the number of observations
dropped from 65 to 26 and the number of athletes included in the analysis
decreased from 15 to 9.

@@

\section{Discussion}\label{sec:Discussion}
Since 1985 Seiko Holding Corporations has served every World Athletics 
Championships as the official timer \citep{wa2022seiko}.  Since 1985, the 
technology and the ability to accurately predict measurements: long jump 
distances, false starts, total time, reaction time, etc. have all dramatically 
improved.  Seiko did not start tracking reaction time as an official measurement
until 1999 in the men's 110 meter hurdles and women's 100 meter hurdles. 
Seiko regularly updates their equipment so that they provide cutting edge
technology to the World Track and Field Championships, the highest stakes in the 
world of running outside of the Olympics. Seiko's technology for detecting 
reaction times relies on the pressure that athletes exert on the plate when they 
push off.  Their systems measure the time differential between when the ``gun" 
goes off to start the race and when the pressure changes \citep{wa2022seiko}.  
If an athlete has a reaction time under 0.1 seconds, it is deemed a false start 
as it is considered that no athlete can react so quickly \citep{Seiko-Timing}.  
Thus, a reaction time of for example 0.05, suggests that the athlete predicted 
the gun and it was luck that caused their abnormally high time.  At the World 
Championships level, World Athletics and Seiko want to remove that element of 
luck and thus impose the 0.1 second barrier. In 2013, Seiko upgraded their 
timing equipment for sprinting events (110 meter hurdles falls under this 
category) \citep{wa2013backtage}. Since 1999, the slowest reaction times in the 
men's 110-meter hurdle were recorded in 2013.  That is not to say that the higher 
times were caused by Seiko's equipment, but the two may be related.  It is worth 
noting that prior to 2022, Seiko again upgraded its technology but for its jump 
management system. 
\eds{I would argue that this paragraph should go in the discussion as there are 
no results presented here}
\of{I agree so I moved it here. I think it also could fit into the top of the
data section as in section~\ref{sec:Data}. Thoughts?}

@@

One conclusion from the analysis in \citet{zhang2021correlation} was that 
athletes' reaction times improved from 2011 to 2019.  However this study does 
not consider data from 1999 to 2009, and over this timespan athletes get slower 
and then faster.  Figure~\ref{fig:Boxplots} shows that for the men's pooled data,
the year with the fastest median reaction time before 2022 was 1999.  Thus, this 
reject's Zhang et al.'s conclusion that athletes are getting faster because they 
were fast over twenty years ago. One possible explanation for why 2011 had such
high variation was because of the IAAF rule discussed earlier that increased the
penalty of a false start to an instant disqualification \citep{iaaf2009falsestart}.
Athletes may have been overly cautious to false start and may have consciously
or sub-consciously reacted slower as a result.  Instead, a more appropriate 
conclusion may be that there is a lot of variability in reaction times and that
the context behind the races matters.

It is worth noting that in the semifinals of the 2022 World Track and
Field Championships, Allen's reaction time was 0.101 which is only 0.001 above
the legal limit. What this suggests is that Allen may have strong reflexes
and be able to react extremely well to the sound of the gun. It seems unlikely 
that Allen would be able to correctly predict the gun with such precision two 
times in a row, which is the exact reason that the 0.1 second reaction time 
disqualification barrier was
imposed in the first place. Thus having a quick reaction time multiple times
may indicate skill rather than luck. Devon 
Allen is a very good athlete, as evidenced by his football career at the 
University of Oregon
and making it onto the Philadelphia Eagles practice squad this past year 
\citep{hurley2022eagles}. The 0.099 reaction time may not have been 
a product of Devon Allen predicting the start of the race but rather a 
combination of a quick reaction and a possibly faulty sensor.

@@

This paper does not consider data from the women's 100 meter hurdle, and although
that was initially because of concerns over reaction time differences in men and
women, exploratory analysis showed that 2022 was at best only
slightly faster relative to other championships going back to 2003 
\eds{check year}. \eds{this seems to contradict the Robert Johnson conclusion in 
the intro, unless he only considered sprint and not hurdles?} Likewise,
the rank-based comparisons resulted in p values that were not significant.  The
lack of evidence from the women's data makes it harder to conclude 
that the sensors used at the 2022 championships were faulty, although it is
unknown if the sensors for the men's 110 meter hurdle and women's 100 meter
hurdle were the same.

It is confusing as to why the reaction time barrier for men's and women's data
is the same despite numerous articles such as \citep[e.g.,][]{lipps2011implications, 
babicc2009reaction, panoutsakopoulos2020gender} stating differently.  World
Athletics is ignoring differences in gender by setting the reaction time barrier
to be equal. If 0.1 seconds is considered a fair time for women, then 0.1 seconds
could not be considered fair for men, as the probability of observing reaction
times lower than 0.1 would be elevated.

>>>>

The dataset of the four United States athletes 
is too small to perform any meaningful statistical analysis, but
the data was expanded to include reaction times for athletes who competed at 
national competitions (many of which were held from May-July of 2022). Once the 
data was collected, each athlete was designated as a cluster and the stage of
competition (national or international) was specified for each competition 
within the cluster (athlete).  We chose to include preliminary rounds in this
section as well as in \ref{sec:data2019} in order to increase the size of our
data.  Without data from the preliminary rounds, the number of observations
dropped from 65 to 26 and the number of athletes included in the analysis
decreased from 15 to 9.  Likewise

@@

\section{Discussion}\label{sec:Discussion}
One conclusion from the analysis in \citet{zhang2021correlation} was that 
athletes' reaction times improved from 2011 to 2019.  However this study does 
not consider data from 1999 to 2009, and over this timespan athletes get slower 
and then faster.  Figure~\ref{fig:Boxplots} shows that for the men's pooled data,
the year with the fastest median reaction time before 2022 was 1999.  Thus, this 
reject's Zhang et al.'s conclusion that athletes are getting faster because they 
were fast over twenty years ago. One possible explanation for why 2011 had such
high variation was because of the IAAF rule discussed earlier that increased the
penalty of a false start to an instant disqualification \citep{iaaf2009falsestart}.
Athletes may have been overly cautious to false start and may have consciously
or sub-consciously reacted slower as a result.  Instead, a more appropriate 
conclusion may be that there is a lot of variability in reaction times and that
the context of the races matters.

This paper does not consider data from the women's 100 meter hurdle, and although
that was initially because of concerns over reaction time differences in men and
women, exploratory analysis showed that 2022 was at best only
slightly faster relative to other championships going back to 2003 
\eds{check year}. \eds{this seems to contradict the Robert Johnson conclusion in 
the intro, unless he only considered sprint and not hurdles?} \of{Johnson's
results considered the mens 100m, mens 110m hurdles, womens 100m but not the 
womens 100m hurdles.} Likewise, the rank-based comparisons resulted in p values
that were not significant.  The lack of evidence from the women's data makes it 
harder to conclude that the sensors used at the 2022 championships were faulty, 
although it is unknown if the sensors for the men's 110 meter hurdle and women's
100 meter hurdle were the same.

It is confusing as to why the reaction time barrier for men's and women's data
is the same despite numerous articles such as \citep[e.g.,][]{lipps2011implications, 
babicc2009reaction, panoutsakopoulos2020gender} stating differently.  World
Athletics is ignoring differences in gender by setting the reaction time barrier
to be equal. If 0.1 seconds is considered a fair time for women, then 0.1 seconds
could not be considered fair for men, as the probability of observing reaction
times lower than 0.1 would be elevated.  \citet{brosnan2017effects} advocated
for gender based reaction time barriers which seems logical if there are biological
differences in reaction times between men and women.

Since 1985 Seiko Holding Corporations has served every World Athletics 
Championships as the official timer \citep{wa2022seiko}.  Reaction times were
not officially measured until 1999 in the men's 110 meter hurdles and women's
100 meter hurdles. Seiko's technology for detecting 
reaction times relies on the pressure that athletes exert on the plate when they 
push off.  Their systems measure the time differential between when the ``gun" 
goes off to start the race and when the pressure changes \citep{wa2022seiko}.  
If an athlete has a reaction time under 0.1 seconds, it is deemed a false start 
as it is considered that no athlete can react so quickly \citep{Seiko-Timing}.  
Thus, a reaction time of for example 0.05, suggests that the athlete predicted 
the gun and it was luck that caused their abnormally high time.  At the World 
Championships level, World Athletics want to remove that element of 
luck and thus impose the 0.1 second barrier.
\eds{I would argue that this paragraph should go in the discussion as there are 
no results presented here}
\of{I agree so I moved it here. I think it also could fit into the top of the
data section as in section~\ref{sec:Data}. Thoughts?}

@@

It is worth noting that in the semifinals of the 2022 World Track and
Field Championships, Allen's reaction time was 0.101 which is only 0.001 above
the legal limit. What this suggests is that Allen may have strong reflexes
and be able to react extremely well to the sound of the gun. It seems unlikely 
that Allen would be able to correctly predict the gun with such precision two 
times in a row, which is the exact reason that the 0.1 second reaction time 
disqualification barrier was imposed in the first place. Thus having a quick 
reaction time repeatedly may indicate skill rather than luck. Devon Allen is
an exceptional athlete, as evidenced by his football career at the University of
Oregon and making it onto the Philadelphia Eagles practice squad in 2022
\citep{hurley2022eagles}.  \citet{komi2009iaaf}'s paper showed that athletes are
capable of reacting in less than 0.1 seconds, and if this is the case, it is not
unreasonable to conclude that an outstanding athlete like Allen is capable of
reacting below the 0.1 second barrier.

The results from the GLMM are described in section~\ref{subsec:Results_GLMM} and
show that there

@@
